FAIR fission track analysis with geochron@home

### 2025-07-21 geochron@home/manuscript/paper.tex
Started implementing Rex Galbraith suggestions

The two analysts defined their own ROIs, resulting in slightly
different density estimates (Figure~\ref{fig:AvP}a) for 25 apatite
grains.  Let $N_{PV}$ and $N_{AC}$ be the number of spontaneous tracks
counted by PV and AC over areas $A_{PV}$ and $A_{AC}$,
respectively. Then the ratio of the estimated track densities are
given by $\hat{\rho}_{PV} = N_{PV}/A_{PV}$ and $\hat{\rho}_{AC} =
N_{AC}/A_{AC}$. The uncertainties of these density ratios are strongly
correlated due to the significant overlap between the counting areas
chosen by the two analysts:
\begin{equation}
  \frac{\mbox{se}(\hat{\rho}_{PV}/\hat{\rho}_{AC})}{\hat{\rho}_{PV}/\hat{\rho}_{AC}}
  \approx
  \sqrt{
    \frac{1}{N_{PV}} + \frac{1}{N_{AC}} - \frac{2\,N_{0}}{N_{PV}\,N_{AC}}
  }
  \label{eq:NO}
\end{equation}

\noindent where $N_{0}$ is the number of tracks counted by both
observers in the area of overlap ($A_0$, say) between their respective
ROIs.  If there is no overlap between the two counting areas (i.e., if
$A_0=0$), then Equation~\ref{eq:NO} reduces to the conventional
coefficient of variation for the ratio of two independent Poisson
variables \citep{galbraith1990a}.  Conversely, if the counting areas
are identical (i.e., if $A_{PV}=A_{AC}=A_0$), and if the two analysts
produce perfectly consistent counts (i.e. if $N_{PV} = N_{AC} =
N_{0}$), then Equation~\ref{eq:NO} returns a zero value.\medskip

The radial plot of Figure~\ref{fig:AvP}b contains two sets of counts,
with $\sum N_{PV} = 686$ and $\sum N_{AC} = 679$, and a weighted mean
track density ratio $\hat{\rho}_{PV}/\hat{\rho}_{PV} = 0.94$ with
relative standard error 0.013. PV counted 600 tracks in those common
areas $A_0$, of which 549 were also counted by AC. Conversely, AC
counted 622 tracks of which 549 were also counted by PV. The ratio of
the two analysts' track density estimates based just on the common
area is therefore 600/622 = 0.965 with relative standard error 0.016,
calculated from Equation~\ref{eq:NO}. This is close to the weighted
mean ratio of 0.94 (Figure~\ref{fig:AvP}b), and is slightly nearer to
1. It indicates that PV under-counts the Mount Dromedary apatite by
3.5\% relative to AC.  The existence of `observer bias' is well
documented \citep{tamer2025}.  It is one of the reasons why fission
track analysis is often done relative to age standards: observer bias
does not have to be a problem provided that it is consistent between
grains, and between samples.\medskip

%\begin{figure}[!ht]
{ \centering \includegraphics[width=12cm]{AvP.pdf}
  \captionof{figure}{Comparison of fission track data for Mount
    Dromedary apatite by two analysts (PV = Pieter Vermeesch and AC =
    Andrew Carter). a) the track counts and ROIs of both users for
    grain 4648, shown in blue (PV) and red (AC); b) radial plot for
    the ratio of the track densities of both users for all 25 grains
    analysed by the two analysts, using Equation~\ref{eq:NO}.
    \medskip}
  \label{fig:AvP}
}%\end{figure}

@@

Figure~\ref{fig:radialcrowd} presents the results of a crowd-sourcing
experiment that was carried out at UCL as part of an undergraduate
course in isotope geology. 68~students were asked to create an account
on the geochron@home server by selecting a unique username and
password. After spending a few minutes to complete the compulsory
tutorial (Section~\ref{sec:tutorial}), the students were asked to
count fission tracks in randomly assigned image stacks of Mount
Dromedary apatite. Each student was required to analyse at least 15
grains, using pre-defined ROIs. In a matter of hours, the students
amassed a dataset of 37,331 fission track counts in 34 grains. This
large dataset was parsed into separate data files --one for each
student-- which they processed during an assessed practical
exercise.\medskip

The raw counts are shown as a series of box plots in
Figure~\ref{fig:radialcrowd}a for the same selection of 25 grains that
were analysed by AC and PV in Section~\ref{sec:private}. The box plots
exhibit significant variability between the grains and between the
students. The inter-grain variability is expected. It reflects the
different sizes of the ROIs, and the variable U-content of the
grains. The inter-student variability is caused by observer bias. This
phenomenon was previously discussed in
Section~\ref{sec:private}. However, the bias is much stronger for the
crowd-sourced dataset than it was for the two `expert' analysts of
Section~\ref{sec:private}. The majority of the student analysts
undercounted the fission tracks by up to a factor of five compared to
the `expert' analyst (PV, shown in blue). The poor consistency of the
results reflects the very limited training received by the
students.\medskip

Grains 23 and 25 have greater than average track densities, and are
particularly dispersed. Plotting the 37 paired counts for each user on
a scatter plot reveals a strong correlation
(Figure~\ref{fig:radialcrowd}b). This means that students who over- or
undercount the first grain, also tend to over- or undercount the
second grain. Even though most students count fewer fission tracks
than PV, the ratio of their counts are similar.
Figure~\ref{fig:radialcrowd}c compares grains 23 and 25 on a radial
plot, where
\begin{equation}
  \frac{\mbox{se}(N_1/N_2)}{N_1/N_2}
  \approx
  \sqrt{
    \frac{1}{N_1} + \frac{1}{N_2}
  }
  \label{eq:N12}
\end{equation}

The tight clustering of the data within the 2-se band confirms that
the pairwise ratios of the counts for each analyst are more consistent
than the raw counts of Figure~\ref{fig:crowdsourcing}a.  Despite the
tremendous scatter of the raw track counts obtained by the citizen
scientists, their pairwise ratio estimates scatter evenly around the
expert opinion. This suggests that the collective wisdom of the entire
group of students is greater than that of any individual
analyst.\medskip

@@

Figure~\ref{fig:crowdsourcing} also shows that some students counted
the tails of the fission tracks rather than their etch pits, despite
being told the opposite in the tutorial. Fixing this issue will
require some improvements to the tutorial pages
(Section~\ref{sec:outlook}).

@@

\begin{enumerate}
\item Interactive tutorial pages. To ensure that novice users do not
  count the tails but the etch pits of fission tracks, we will add a
  `quiz' to the tutorial pages. Only users who click on the `correct'
  features in an unlabelled set of images will be allowed to count new
  samples.
\item Length measurements. The geochron@home archive already
  contains horizontally confined fission track measurements
  \citep{tamer2025}.  However, these results must be generated
  externally (e.g., using Fission Track Studio) and uploaded via a
  \texttt{.json} file.  The virtual microscope curently lacks the
  functionality to generate length data within
  geochron@home. This functionality will be added in a future
  update.
\item Dpar and Dper. Etch pits are currently stored as simple sets of
  x- and y-coordinates.  In reality, etch pits have a finite length
  (`Dpar') and width (`Dper'), which serve as useful indicators for
  the horizontal etch rates along the c-axis and parallel to it
  \citep{donelick1993}. Functions will be added measure and visualise
  these type of data in geochron@home.
\item Machine learning. Data science is experiencing an artificial
  intelligence revolution that has already started to transform the
  fission track method \citep{nachtergaele2020}.  Convolutional neural
  networks must be trained with example data. geochron@home
  is ideally suited for this task. Section~\ref{sec:crowdsourcing}
  showed how the collective wisdom of multiple fission track analysts
  can label fission track images with probabilities rather than
  counts. This data format is close to the form in which data are
  treated within an AI algorithm.\medskip

>>>>

Let $N_1$ and $N_2$ be the numbers of spontaneous tracks counted by PV
and AC over areas $A_1$ and $A_2$, respectively. Then their estimated
track densities are give by $\hat{\rho}_1 = N_1/A_1$ and $\hat{\rho}_2
= N_2/A_2$. In this situation when the two areas overlap, $N_1$ and
$N_2$ cannot be treated as independent Poisson counts because the two
analysts will count some of the same tracks. It can be shown that,
under simple (ideal) assumptions, the uncertainty of the ratio of the
estimated track densities is given approximately by
\begin{equation}
  \frac{\mbox{se}(\hat{\rho}_1/\hat{\rho}_2)}{\hat{\rho}_1/\hat{\rho}_2} \approx
  \sqrt{ \frac{1}{N_1} + \frac{1}{N_2} - \frac{2\,N_0}{N_1\,N_2} }
  \label{eq:NO}
\end{equation}

\noindent where $N_0$ is the number of tracks counted by both
observers in the area of overlap ($A_0$, say) between their respective
ROIs.\medskip

The radial plot of Figure~\ref{fig:AvP}b contains two sets of counts,
with $\sum N_{1} = 686$ and $\sum N_{2} = 679$, and a weighted mean
track density ratio $\hat{\rho}_{1}/\hat{\rho}_{1} = 0.94$ with
relative standard error 0.013. PV counted 600 tracks in those common
areas $A_0$, of which 549 were also counted by AC. Conversely, AC
counted 622 tracks of which 549 were also counted by PV. The ratio of
the two analysts' track density estimates based just on the common
area is therefore 600/622 = 0.965 with relative standard error 0.016,
calculated from Equation~\ref{eq:NO}. This is close to the weighted
mean ratio of 0.94 (Figure~\ref{fig:AvP}b), and is slightly nearer to
1. It indicates that PV under-counts the Mount Dromedary apatite by
3.5\% relative to AC.  The existence of `observer bias' is well
documented \citep{tamer2025}.  It is one of the reasons why fission
track analysis is often done relative to age standards: observer bias
does not have to be a problem provided that it is consistent between
grains, and between samples.\medskip

%\begin{figure}[!ht]
{ \centering \includegraphics[width=\linewidth]{AvP.pdf}
  \captionof{figure}{Comparison of fission track data for Mount
    Dromedary apatite by two analysts (PV = Pieter Vermeesch and AC =
    Andrew Carter). a) the track counts and ROIs of both users for
    grain 4648, shown in blue (PV) and red (AC); b) comparison of the
    track densities for PV and AC for all 25 grains analysed by the
    two analysts with a reference line for
    $\hat{\rho}_\text{PV}=\hat{\rho}_\text{AC}$; c) radial plot of the
    same data, using Equation~\ref{eq:NO}.}
  \label{fig:AvP}
}%\end{figure}

@@

Figure~\ref{fig:radialcrowd}a shows box plots and summary statistics
for the raw counts made by 68 students (two of the students were
unable to complete the assignment) for the same selection of 25 grains
that were analysed by AC and PV in Section~\ref{sec:private}. Everyone
counted tracks in the same ROIs, and counts made by PV for those ROIs
have been added as blue dots.  Variation between the mean values is to
be expected because of the differing numbers of tracks due to varying
areas and uranium contents of the grains. But the variation between
counts within grains (shown by the standard deviations) is due
entirely to differences between the students' recognising and counting
exactly the same tracks. These standard deviations increase with the
mean number of tracks and are considerable in size, being on average
about 40\% of the mean.  Also, the vast majority of students counted
fewer tracks than PV did, often many fewer. In 21 of the 25 grains,
PV's count is above the upper quartile of the students', and on 2
occasions someone counted no tracks at all.  With expert trained
counters one would expect much smaller differences between counts.
Nevertheless there are many repeated counts also.\medskip

Figure~\ref{fig:radialcrowd}b shows a scatter plot of pairs of counts
for two grains made by 33 students who counted both of them. They
average around 23 and 30 tracks per grain compared with PV's pair of
44 and 63, which is highlighted in blue. There is a positive
correlation between the pairs, consistent with systematic observer
effects (i.e., people who count a low value in one grain tend to count
similarly low in the other, and vice versa).  Comparing other pairs of
grains shows similar results, with correlations varying between 0.3
and 0.9. In all cases we have looked at, the residual variation, after
allowing for systematic differences between students, is consistent
with Poisson variation.  Figure~\ref{fig:radialcrowd}c shows a radial
plot of the ratios of the counts in Panel b, using the precisions
given by
\begin{equation}
  \frac{\mbox{se}(N_1/N_2)}{N_1/N_2}
  \approx
  \sqrt{
    \frac{1}{N_1} + \frac{1}{N_2}
  }
  \label{eq:N12}
\end{equation}

The context here is rather different from that in
Figure~\ref{fig:AvP}b, where we were comparing the ratios of counts
made by 2 analysts on each of 25 grains. Here we are comparing the
ratios of counts for the same 2 grains made by each of 33
analysts. There is no theoretical justification for
Equation~\ref{eq:N12} here, but it still provides a useful benchmark
for assessing variation between the ratios. In
Figure~\ref{fig:radialcrowd}c, that variation is consistent with
Poisson variation (as it is for most other pairs of grains) but with
expert counters one would expect less variation than that. However, in
spite of the under-counting by the students, the pooled the students'
ratio of 0.76 is quite close to PV's ratio of 0.70. This happens quite
often in this data set, but not always, and there are some pairs of
grains where the students' pooled ratio differs somewhat from
PV's.\medskip

@@

Figure~\ref{fig:crowdsourcing} also shows that some students counted
the tails of the fission tracks rather than their etch pits, despite
being told the opposite in the tutorial. Fixing this issue will
require some improvements to the tutorial pages
(Section~\ref{sec:outlook}).\medskip

@@

\begin{enumerate}
\item Interactive tutorial pages. To ensure that novice users do not
  count the tails but the etch pits of fission tracks, we will add a
  `quiz' to the tutorial pages. Only users who click on the `correct'
  features in an unlabelled set of images will be allowed to count new
  samples. The pilot data of the crowd-sourcing experiment show that,
  despite the large scatter of the raw track counts obtained by the
  citizen scientists, their pairwise ratio estimates scatter evenly
  around the expert opinion. This suggests that the collective wisdom
  of the entire group of students is greater than that of the
  individual analysts. When applied to a group that has achieved a
  good level of training and experience, crowd sourcing should improve
  the precision and accuracy of fission track data.
\item Length measurements. The geochron@home archive already
  contains horizontally confined fission track measurements
  \citep{tamer2025}.  However, these results must be generated
  externally (e.g., using Fission Track Studio) and uploaded via a
  \texttt{.json} file.  The virtual microscope curently lacks the
  functionality to generate length data within
  geochron@home. This functionality will be added in a future
  update.
\item Dpar and Dper. Etch pits are currently stored as simple sets of
  x- and y-coordinates.  In reality, etch pits have a finite length
  (`Dpar') and width (`Dper'), which serve as useful indicators for
  the horizontal etch rates along the c-axis and parallel to it
  \citep{donelick1993}. Functions will be added measure and visualise
  these type of data in geochron@home.
\item Machine learning. Data science is experiencing an artificial
  intelligence revolution that has already started to transform the
  fission track method \citep{nachtergaele2020}.  Convolutional neural
  networks must be trained with example data. geochron@home
  is ideally suited for this task. Section~\ref{sec:crowdsourcing}
  showed how the collective wisdom of multiple fission track analysts
  can label fission track images with probabilities rather than
  counts. This data format is close to the form in which data are
  treated within an AI algorithm.\medskip
